A semi-autonomic framework for developing Machine Learning-based applications using Fairness metrics

### 2023-10-20 texts/article/article-eng.tex
Thesis version 1.1.0 and article version 1.1

%% Tables
%%\usepackage{booktabs}

@@

\title{A semi-autonomic solution for developing Machine Learning-based applications using Fairness metrics}
%\titlerunning{Semi-autonomic Workflow for ML applications using \textit{Fairness} metrics}

%\author{Thales Eduardo Nazatto\textsuperscript{1}, Cecília Mary Fischer Rubira\textsuperscript{2}, Leonardo Montecchi\textsuperscript{3}}
%\authorrunning{Thales Eduardo Nazatto, Cecília Mary Fischer Rubira, Leonardo Montecchi}
%\author{
%	\IEEEauthorblockN{Thales Eduardo Nazatto}
%	\IEEEauthorblockA{
%		UNICAMP, Campinas, Brazil \\
%		tenazatto@gmail.com}
%	\and
%	\IEEEauthorblockN{Cecília Mary Fischer Rubira}
%	\IEEEauthorblockA{
%		UNICAMP, Campinas, Brazil \\
%		cmrubira@ic.unicamp.br}
%	\and
%	\IEEEauthorblockN{Leonardo Montecchi}
%	\IEEEauthorblockA{
%		NTNU, Trondheim, Norway \\
%		leonardo.montecchi@ntnu.no}
%}

\author{Thales Eduardo Nazatto}
\email{tenazatto@gmail.com}
\affiliation{%
  \institution{University of Campinas}
  \streetaddress{P.O. Box 1212}
  \city{Campinas}
  \state{São Paulo}
  \country{Brazil}
  \postcode{43017-6221}
}

@@

\author{Leonardo Montecchi}
\email{leonardo.montecchi@ntnu.no}
\affiliation{%
  \institution{Norwegian University of Science and Technology}
  \city{Trondheim}
  \state{Trøndelag}
  \country{Norway}
}

%\date{
%	\textsuperscript{1} UNICAMP, Campinas, Brazil, tenazatto@gmail.com
%	\\
%	\textsuperscript{2} UNICAMP, Campinas, Brazil, cmrubira@ic.unicamp.br
%	\\
%	\textsuperscript{3} NTNU, Trondheim, Norway, leonardo.montecchi@ntnu.no\\[2ex]
%	\today
%}

%\institute{
%	\textsuperscript{1} UNICAMP, Campinas, Brazil, tenazatto@gmail.com
%	\\
%	\textsuperscript{2} UNICAMP, Campinas, Brazil, cmrubira@ic.unicamp.br
%	\\
%	\textsuperscript{3} NTNU, Trondheim, Norway, leonardo.montecchi@ntnu.no\\[2ex]
%}

%\settopmatter{printacmref=false}
%\setcopyright{none}
%\renewcommand\footnotetextcopyrightpermission[1]{}
\setcopyright{acmcopyright} % if you give the rights to ACM
\acmDOI{11.1111/1111111.1111111} % DOI - Insert your DOI below...
\acmISBN{111-1-1111-1111-1} % ISBN - Insert your conference/workshop's ISBN below...
\acmYear{2023} % Insert Publication year
\copyrightyear{2023} % Insert Copyright year (typically the same as above)
\acmPrice{0.00}
\acmConference[LADC 2023]{12th Latin-American Symposium on Dependable and Secure Computing}{October 16-20 2023}{La Paz, Bolívia}

@@

\begin{abstract}
Machine Learning (ML) is increasingly used with big data to make faster and more assertive decisions, but initial metrics to evaluate algorithms have proven limited in measuring biases that reflect society in an unexpected way. To address this, new Fairness algorithms and metrics have been created to balance discriminated groups, but it increases the complexity for Data Scientists to develop the best models. This article focuses on Machine Learning solutions from a Software Engineering point of view, presenting a Workflow that uses the MAPE-K architecture to determine which algorithm has the best balance, without having to test many techniques. Several case studies were carried out to verify the feasibility of MAPE-K in solving problems, and if the solution can be easily extended to allow the use of more algorithms. This can help ML Specialists to do Software Engineering studies on applications with responsible use of data.
\end{abstract}

@@

\keywords{Automated Machine Learning, AI Ethics, Social Responsibility in AI, Autonomic Loop, Fairness metrics, MAPE-K}

@@

Artificial Intelligence (AI) and Machine Learning (ML) techniques have been used for a long time in the field of Computer Science. Areas such as robotics and games are great examples, given their need to automate behaviors that would be considered trivial for a human being. However, in recent years there has been a growth in the use of these technologies in traditional applications, mainly due to the large amount of data processed daily by companies and the amount of processing available at low costs. Different profiles can be drawn from these data and using AI solutions generate more assertive decision making in order to improve the user experience and correct problems. However, a common side effect in these solutions is the exposure of biases that, although seen as unintentional by developers due to the possibility of being an \textit{outlier} in the trained model, reflect open prejudices of today's society. Biased data entries result in an algorithm that performs discriminations in its classification~\citep{Buolamwini_2018}, and since the metrics used to measure the quality of final model are generally based on accuracy, precision and recall, social contexts and discrimination are not easily perceived by such metrics.

Due to this problem, it is possible to establish different metrics to determine how well the model is prepared for sensitive data~\citep{Begley_2021}, a term that is known as \textit{Fairness}. With the evolution of academic research on this concept, new methods were defined to reduce the biases present in datasets, such as \textit{Reweighing}~\citep{Kamiran_2011}, \textit{Adversarial Debiasing}~\citep{Zhang_2018} and \textit {Reject Option Classification}~\citep{Kamiran_2012}. Consequently, there is an improvement in these metrics, but it may disfavor metrics that are already widely used as a guarantee of an effective model developed with Machine Learning techniques.

For this particular context, a method for developing Machine Learning applications was developed, which is handled by Data Scientists in a semi-autonomous way, focused on two main objectives:

\begin{itemize}
\item Make the creation of fair and reliable models by automating the choice of algorithms easier, due to increased complexity in choosing algorithms to be used and their execution in the correct steps of the process.
\item Define a balance between metrics for evaluating good models with metrics for evaluating fair models.
\end{itemize}

This structure is divided into 5 modules: Data Engineering, ML Workflow, Autonomic Manager, Backend and Frontend. For Data Engineering, modifications were made to the German Credit Dataset~\citep{ucigerman_2021} and Lendingclub Dataset~\citep{lendingclub_2022} datasets to be suitable for the \textit{workflow} implementation. For the workflow development, the \textit{Pipes-and-Filters} architecture was used. For the Autonomic Manager, the MAPE-K~\citep{IBM_2005} architecture was used to analyze a knowledge base and provide the best configuration background following predetermined rules. For the Frontend and the Backend, its creation was thought of a web application template\footnote{Repository containing the codes for this project: \url{https://anonymous.4open.science/r/FAIR-2EFD}} for evaluation and testing in further studies.

\section{Background}

\subsection{Autonomic Computing}

In 2001, Paul Horn introduced the concept of Autonomic Computing as a possible solution to the growing complexity of systems at the time, where it was predicted that they would become too large and complex for even the most qualified professionals to configure and maintain. This concept qualifies computing systems that can self-manage in relation to the high-level goals given by administrators and is derived from biology, given the great variety and hierarchy of autonomous systems existant in nature and society~\citep{Kephart_2003}.

To Autonomic Computing works as intended, an Autonomic Element~\citep{Abbas_2010} is implemented, a software component that manages parts of the system based on a MAPE-K \textit{loop} (\textit{Monitor, Analyze, Plan, Execute, and Knowledge}). MAPE-K is a concept that constitutes a control \textit{loop}, used to monitor and control one or more Managed Elements. A Managed Element can be hardware such as a printer, software such as a database, another Autonomous Element, or specific functions such as load balancing.

\begin{figure}[H]
\centering
\includegraphics[scale=0.5]{images/MAPE-K.png}
\caption {MAPE-K architecture diagram~\citep{Abbas_2010}.}
\label{fig:MAPEK}
\end{figure}

As illustrated in Figure \ref{fig:MAPEK}, a MAPE-K control \textit{loop} is split as follows:

\begin{itemize}
\item \textbf{Monitor:} Responsible for monitoring managed resources and collecting, aggregating and filtering data. Monitoring is done through one Sensor or more.
\item \textbf{Analyze:} Analyzes the data reported by the monitor part. The analysis aims to understand what the current state of the system is and if there are measures to be taken.
\item \textbf{Plan:} An action plan is prepared in the basis of the analysis' results. The plan is a series of measures that will move the system from its current state to a desired state.
\item \textbf{Execute:} The plan is executed and controlled. One Effector or more perform the planned actions on the resource.
\item \textbf{Knowledge:} The knowledge base is central and accessible by all parts of the \textit{loop}. Separated from collected and analyzed data, it contains additional knowledge such as architectural models, goal models, policies and change plans.
\end{itemize}

\subsection{Machine Learning}
Machine Learning can be defined as “the practice of using algorithms to collect data, learn from it, and then make a determination or prediction about something in the world. Instead of doing code manually to complete a particular task, the machine is `trained` using large amount of data and algorithms that give it the ability to learn how to perform tasks.

The method for training a machine are composed by the following steps:

\begin{itemize}
\item Data collection
\item Data cleaning and refinement
\item Data preparation and split into training and test sets
\item Algorithm training and learning using the training set
\item Metric evaluation using the test set
\end{itemize}

A Machine Learning algorithm can be classified into three basic categories:

\begin{itemize}
\item \textbf{Supervised learning}: Training is performed using labeled examples, such as an input in which the desired output is known. Through methods such as classification, regression, and \textit{gradient boosting}, supervised learning uses patterns to predict the values of labels in additional unlabeled data. Supervised learning is commonly employed in applications where historical data predict likely future events.
\item \textbf{Unsupervised Learning}: Used on data that does not have historical labels. The “right answer” is not reported to the system, the algorithm must figure out what is being shown. The goal is to explore the data and find some structure within it. Popular techniques include self-organizing maps, proximity mapping, \textit{k-means}, and singular value decomposition. These algorithms are also used to segment text topics, recommend items, and identify outliers in data.
\item \textbf{Reinforcement learning}: The algorithm discovers through trial-and-error testing which actions yield the greatest rewards. This type of learning has three main components: the agent (the learner or decision maker), the environment (everything the agent interacts with) and actions (what the agent can do). The objective is for the agent to choose actions that maximize the expected reward in a given period of time. The agent will reach the goal much faster if it follows a good policy, so the focus of reinforcement learning is to discover the best policy.
\end{itemize}

\subsection{\textit{Fairness} in Machine Learning}

\begin{figure*}[h]
\centering
\includegraphics[scale=0.4]{images/backend-frontend-ml-eng.jpg}
\caption {Architecture components of the proposed solution}
\label{fig:BackendFrontendML}
\end{figure*}

It is possible to describe the concept of \textit{Fairness} in the context of supervised learning, where a model $f$ can predict a set of results $y$ from a set of \textit{features} $x$, avoiding unfair discrimination against a protected attribute $a$. It is allowed, but not required, for $a$ to be a component of $x$~\citep{Begley_2021}. In other words, a fair ML model is one where the correlation of its result is low in relation to input data considered sensitive to discrimination. Machine Learning algorithms are not able to differentiate social contexts, where a more efficient result according to the available data can amplify social inequalities and make decisions unfairly~\citep{Mehrabi_2021}. These sensitive data, such as race, sex, age and height, are considered protected attributes, which need to be classified and processed before the execution of a Machine Learning algorithm, which will determine how the algorithm will behave and, consequently, will affect the metrics~\citep{Mougan_2022}. The groups of data from these protected attributes are considered protected groups, which can be divided into two groups: the privileged group, which has advantages in the problem's context, and the non-privileged group, which has disadvantages in the problem's context and, therefore, subject to discrimination.

Fairness metrics differ from the metrics used for model evaluation, which have the purpose of checking whether a model has reliable predictions or not. While the more traditional metrics evaluate the performance of the model, data as a whole and general results, Fairness metrics evaluate whether the general results are also reflected in specific groups, to verify that there is no disparity or discrimination in the proposed results. Examples of metrics used for this are Statistical Parity Difference, or discrimination~\citep{Zemel_2013}, Equal Opportunity Difference~\citep{Biswas_2020}, Average Odds Difference~\citep{Biswas_2020}, Disparate Impact~\citep{Biswas_2020} and Theil Index~\citep{Speicher_2018}. Almost all of these metrics needs a value close to 0 to be considered fair. The only exception is Disparate Impact which needs a value close to 1 due to the fact this metric is the only one which is calculated as a ratio.

To balance these metrics in a ML model, new algorithms were created to make the learning process more prepared to deal with biased data, which can be classified into three categories:

\begin{itemize}
\item \textbf{Pre-processing:} These algorithms try to remove biased data by transforming it. Some examples are \textit{Disparate Impact Remover}~\citep{Feldman_2015}, \textit{Learning Fair Representations}~\citep{Zemel_2013}, \textit{Reweighing}~\citep{Kamiran_2011} and \textit{Optimized Preprocessing}~\citep{Calmon_2017}
\item \textbf{Processing:} These algorithms try to mitigate discrimination during the training process. Some examples are \textit{Adversarial Debiasing}~\citep{Zhang_2018}, \textit{Exponentiated Gradient Reduction}~\citep{Agarwal_2018}, \textit{Grid Search Reduction}~\citep{Agarwal_2019}, \textit{Meta Fair Classifier}~\citep{Celis_2019}, \textit{Prejudice Remover}~\citep{Kamishima_2012} and \textit{Rich Subgroup Fairness}~\citep{Kearns_2018}
\item \textbf{Post-processing:} These algorithms change the predictions made by an already trained model to make it fair. Some examples are \textit{Equalized Odds}~\citep{Hardt_2016}, \textit{Calibrated Equalized Odds}~\citep{Pleiss_2017} and \textit{Reject Option Classification}~\citep{Kamiran_2012}.
\end{itemize}

\section{The Proposed Solution}

@@

%\begin{figure*}[h]
%\centering
%\includegraphics[scale=0.35]{images/Diagrama_Sequencia-eng.jpg}
%\caption {System's sequence diagram}
%\label{fig:DiagramaSequencia}
%\end{figure*}
\begin{sidewaysfigure*}
\clearpage
\includegraphics[scale=0.425]{images/Diagrama_Sequencia-eng.jpg}
\caption {System's sequence diagram}
\label{fig:DiagramaSequencia}
\end{sidewaysfigure*}

The solution architecture is divided into 5 modules:

\begin{itemize}
    \item {\textbf{Data Engineering:}} Module with the objective of executing data transformation and cleaning processes.
	\item {\textbf{ML Workflow:}} Module that executes a Pipeline of an automated ML application, with stages of data preparation (Pre-processing), training (Processing) and result evaluation (Post-processing) of the final model generation.
     \item {\textbf{Autonomic Manager:}} Module containing a MAPE-K \textit{loop} that controls the Workflow as a Managed Element to automate all its steps, in order to avoid manual executions depending on the algorithm and/or dataset used on the stages.
     \item {\textbf{User Interface (Frontend):}} Module which goal is to provide a simpler and more intuitive user experience for the Data Scientist in order to configure and start the Autonomic Manager.
	\item {\textbf{Backend:}} Module created with the objective of connecting the Frontend to the Autonomic Manager.
\end{itemize}

The integration between these modules is illustrated in Figure \ref{fig:BackendFrontendML}. The Data Engineering module is used by the Data Engineer, who would return a dataset with the necessary treatments for execution within the supervised learning workflow. The Frontend module would be used by the Data Scientist, who would determine the settings needed to choose the configuration background depending on the problem's context.

The process that takes place, from the User Interface interaction, passing through the execution of workflow and finishing at obtaining the results, can be visualized in the Sequence Diagram present in Figure \ref{fig:DiagramaSequencia}. In this process, while the Data Engineer handles the dataset with transformation and cleaning processes presented in the Data Engineering module and send the modified datasets to integrate and use into ML Workflow, the Data Scientist, via the application's User Interface, choose the dataset and demand an Autonomic Manager execution. The Autonomic Manager queries all the executions made by the ML Workflow that used this dataset and are saved into the Knowledge Repository, do the analysis making some calculations that will be detailed in the next sections, and do a loop verifying the configuration of some implemented execution plans and executing them, returning some best execution options and wait the Data Scientist evaluation to choose one option to request a new ML Workflow execution. After the request, the ML Workflow starts the Monitor to collect data during its execution, get the chosen dataset and protected attribute, split into training and test sets, execute the chosen algorithms in the Pre-processing, Processing and Post-processing steps, and calculate the final model metrics. The Monitor save all collected data into the Knowledge Repository, and the Autonomic Manager queries this performed execution data to show to the Data Scientist.

\subsection{Monitor and Analysis Component behavior}

To guarantee the workflow autonomy, the Monitor component collects the data obtained during a execution (metrics, parameters and executed steps) and organizes them for the Analyzer to make the data analysis process. In this process, a calculation of weights based on specific metrics is performed to consolidate all metrics and simplify planning strategies. Metrics are divided into two groups (Performance metrics and Fairness metrics), and within this group you can place as many metrics as necessary, as long as the context of each group is respected. Each group is assigned different weights, and each metric in that group is also assigned different weights. The reason these weights exist in the calculation is to give a measure of the problem's context, according to a prior analysis by the Data Scientist and Domain Specialist.

First, it is necessary to standardize the Fairness metrics $m_{F_i}$ to $m'_{F_i}$ in an interval from 0 to 1, as presented in Equation \ref{eqn:normalizationFairness}. This guarantees the result consistency and the correct application of weights, avoiding distortions in the calculation. As Fairness metrics can involve both the ratio between two different values, like \textit{Disparate Impact}, and the difference between two different values, like \textit{Equal Opportunity Difference}, different calculations are required to get that range. As the evaluated Performance metrics have the same range from 0 to 1, there is no need for additional treatment for the metrics $m_{P_i}$.

Then, each standardized metric is multiplied by its corresponding weights $w_{P_i}$ and $w_{F_i}$, and a weighted average is taken within the group to assign a score $S_P$ to the Performance metrics group and $S_F$ for the Fairness metrics group, as shown in equation \ref{eqn:groupScores}. To make the score evaluation easier, the scores are multiplied by a factor $X = 1000$ to make the score range is 0 to 1000 and the value is rounded. After such scores are obtained, the overall score $S$ is calculated by multiplying them by their corresponding weights $w_P$ and $w_F$ and arranging the weighted average as shown in equation \ref{eqn:totalScore}.

\begin{gather}
\label{eqn:normalizationFairness}
	m'_{F_i} = 
	\begin{cases}
	1-\lvert m_{F_i} \rvert & \text{if $m_{F_i}$ is subtraction-calculated}\\
	& \text{and $-1 < m_{F_i} < 1$}\\
	0 & \text{if $m_{F_i}$ is subtraction-calculated}\\
	& \text{and $m_{F_i} >= 1$ or $m_{F_i} <= -1$}\\
	1-\lvert \frac{1}{m_{F_i}}-1 \lvert & \text{if $m_{F_i}$ is division-calculated and $m_{F_i} > 1$}\\
	1-\lvert m_{F_i}-1 \lvert & \text{if $m_{F_i}$ is division-calculated and $m_{F_i} <= 1$}\\
	m_{F_i} & \text{in other cases}
	\end{cases}
\end{gather}

\begin{equation}
\label{eqn:groupScores}
	\begin{aligned}
	S_F = \round{X \times \frac{\sum_{i=1}^{n_F} w_{m'_{F_i}} \times m'_{F_i}}{\sum_{i=1}^{n} w_{m'_{F_i}}}}\\
	S_P = \round{X \times \frac{\sum_{i=1}^{n_P} w_{m_{P_i}} \times m_{P_i}}{\sum_{i=1}^{n} w_{m_{P_i}}}}
	\end{aligned}
\end{equation}

\begin{equation}
\label{eqn:totalScore}
	S = \frac{w_F \times S_F + w_P \times S_P}{w_F + w_P}
\end{equation}

To do these calculations, previous executions in the workflow are needed as a prerequisite, so the necessary metrics could be recorded and then the best combinations are determined through the score.

>>>>

%% Tables
\usepackage{booktabs}

\usepackage{comment}

@@

\title{A semi-autonomic framework for developing Machine Learning-based applications using Fairness metrics}
\begin{comment}
\author{Thales Eduardo Nazatto}
\email{tenazatto@gmail.com}
\affiliation{%
  \institution{University of Campinas}
  \streetaddress{P.O. Box 1212}
  \city{Campinas}
  \state{São Paulo}
  \country{Brazil}
  \postcode{43017-6221}
}

@@

\author{Leonardo Montecchi}
\email{leonardo.montecchi@ntnu.no}
\affiliation{%
  \institution{Norwegian University of Science and Technology}
  \city{Trondheim}
  \state{Trøndelag}
  \country{Norway}
}
\end{comment}

%\settopmatter{printacmref=false}
%\setcopyright{none}
%\renewcommand\footnotetextcopyrightpermission[1]{}
\setcopyright{acmcopyright} % if you give the rights to ACM
\acmDOI{xx.xxx/xxx_x} % DOI - Insert your DOI below...
\acmISBN{979-8-4007-0243-3/24/04} % ISBN - Insert your conference/workshop's ISBN below...
\acmYear{2024} % Insert Publication year
\copyrightyear{2024} % Insert Copyright year (typically the same as above)
\acmArticle{4}
\acmPrice{15.00}
\acmConference[SAC'24]{ACM SAC Conference}{April 8-12 2024}{Avila, Spain}

\hyphenation{ex-ten-si-bil-i-ty}
\hyphenation{com-put-ing}
\hyphenation{soft-ware}
\hyphenation{re-spon-si-bil-i-ty}
\hyphenation{Mon-tec-chi}
\hyphenation{ob-jec-tives}
\hyphenation{au-tomat-ing}
\hyphenation{en-gi-neer-ing}
\hyphenation{au-to-mate}
\hyphenation{au-to-mat-ed}
\hyphenation{reweigh-ing}
\hyphenation{re-mover}
\hyphenation{ob-tain-ing}
\hyphenation{ex-po-nen-ti-at-ed}
\hyphenation{im-ple-ment-ed}
\hyphenation{dis-cussed}
\hyphenation{ques-tion-naire}
\hyphenation{clas-si-fi-ca-tion}
\hyphenation{op-ti-mized}
\hyphenation{prob-lems}
\hyphenation{im-ple-men-ta-tion}
\hyphenation{op-por-tu-ni-ty}
\hyphenation{cor-re-spond-ing}
\hyphenation{al-go-rithms}
\hyphenation{be-comes}
\hyphenation{con-tain-ing}
\hyphenation{ex-e-cu-tions}
\hyphenation{met-rics}
\hyphenation{nec-es-sary}
\hyphenation{re-duc-tion}
\hyphenation{au-to-nom-ic}
\hyphenation{pro-posed}
\hyphenation{con-fig-u-ra-tion}
\hyphenation{train-ing}
\hyphenation{dif-fer-ent}
\hyphenation{dif-fer-ence}
\hyphenation{de-vel-op-ers}
\hyphenation{ex-pect-ed}
\hyphenation{mod-i-fi-ca-tion}
\hyphenation{ac-com-mo-date}
\hyphenation{sig-nif-i-cant-ly}
\hyphenation{eval-u-a-tion}
\hyphenation{ar-chi-tec-tur-al}
\hyphenation{aug-men-ta-tion}
\hyphenation{meta-da-ta}
\hyphenation{pro-por-tion-al}
\hyphenation{pro-grammed}
\hyphenation{in-ter-change-able}
\hyphenation{pos-si-bil-i-ties}
\hyphenation{ad-di-tion-al-ly}
\hyphenation{knowl-edge}
\hyphenation{se-quence}
\hyphenation{di-vi-sion}
\hyphenation{pa-ra-me-ter-i-za-tion}

@@

\begin{abstract}
The increasing use of Machine Learning (ML) in digital solutions has created a need for new algorithms and fairness metrics to ensure fairer decisions. However, this has made data scientist analysis more complex, due to the need to balance fairness metrics with traditional evaluation metrics and a greater variety of algorithms. This work presents a semi-autonomous framework to train machine learning models that identifies more optimized configurations across different contexts. The framework uses a module using the \textit{Pipes-and-Filters} architecture with several pre-implemented algorithms and a module using the MAPE-K architecture to evaluate the ideal balance. Several case studies were conducted to determine the feasibility and extensibility of the proposed framework. The results showed that the framework can help data scientists better understand the training process and enable software engineering studies to help train trustworthy machine learning models.
\end{abstract}

@@

\keywords{Machine Learning, Artificial Intelligence Ethics, Social Responsibility in Artificial Intelligence, Autonomic Computing, Fairness Metrics}

@@

Artificial intelligence (AI) and machine learning (ML) techniques have been used for many years in computer science, particularly in areas such as robotics and games, where they can automate tasks and generate insights from large volumes of data. However, these applications can also introduce biases that reflect existing prejudices in today's society. Biased data entries result in an algorithm that performs discriminations in its classification~\citep{Buolamwini_2018}, and evaluation metrics typically evaluate the entire dataset rather than specific groups, so it can be difficult to detect bias in the results.

Due to this problem, it is possible to establish different metrics to determine how well the model is prepared for sensitive data~\citep{Begley_2021}, a term that is known as Fairness. The main goal of Fairness is to ensure that privileged and non-privileged groups, regardless of their protected attributes, are treated equally, as evidenced by similar metrics. Advances in academic research on fairness have led to new methods for reducing biases present in datasets, such as \textit{Reweighing}~\citep{Kamiran_2011}, \textit{Adversarial Debiasing}~\citep{Zhang_2018} and \textit {Reject Option Classification}~\citep{Kamiran_2012}. While these methods have improved fairness metrics, they may disfavor other metrics that are widely used to assess the effectiveness of machine learning models.

For this context, a framework for developing Machine Learning applications was developed~\footnote{Repository containing the code implementation: \url{https://anonymous.4open.science/r/FAIR-2EFD}}, which is handled by Data Scientists in a semi-autonomous way, focused on two main objectives:

\begin{itemize}
\item Facilitate the creation of fair and reliable models by automating the following process: Data preparation, training and evaluation.
\item Establish a balance between Evaluation metrics and Fairness metrics to efficiently obtain fairer models.
\end{itemize}

\begin{figure*}[h]
\centering
\includegraphics[scale=0.285]{images/backend-frontend-ml-eng.jpg}
\caption {Architecture components of the proposed framework}
\label{fig:BackendFrontendML}
\end{figure*}

This structure is divided into 4 modules: Data Engineering, ML Module, Autonomic Manager, and User Interface. The Data Engineering module prepares the data for the ML Module by modifying the German Credit Dataset~\citep{ucigerman_2021} and Lendingclub Dataset~\citep{lendingclub_2022} datasets. The ML Module is developed using the \textit{Pipes-and-Filters} architecture. For the Autonomic Manager, the MAPE-K~\citep{IBM_2005} architecture analyzes a knowledge repository to identify the best architectural configuration for the ML Module. The User Interface is a web application template.

\section{The Proposed Framework}

@@

%\begin{figure*}[h]
%\centering
%\includegraphics[scale=0.35]{images/Diagrama_Sequencia-eng.jpg}
%\caption {System's sequence diagram}
%\label{fig:DiagramaSequencia}
%\end{figure*}
\begin{sidewaysfigure*}
\clearpage
\includegraphics[scale=0.395]{images/Diagrama_Sequencia-eng.jpg}
\caption {Sequence diagram with help of the Autonomic Manager}
\label{fig:DiagramaSequencia}
\end{sidewaysfigure*}

The framework architecture is divided into 4 modules:

\begin{itemize}
    \item {\textbf{Data Engineering:}} Module with the objective of executing data transformation and cleaning processes.
	\item {\textbf{ML Module:}} Module that executes a Pipeline of an automated ML application, with stages of data preparation (Pre-pro\-cess\-ing), training (Processing) and result evaluation (Post-processing) of the final model generation.
     \item {\textbf{Autonomic Manager:}} Module containing a MAPE-K control \textit{loop} that controls the ML Module as a Managed Element to automate all its steps.
     \item {\textbf{User Interface):}} Module which goal is to provide a simpler and more intuitive user experience for the Data Scientist in order to configure and start the Autonomic Manager.
\end{itemize}

The integration between these modules is illustrated in Figure \ref{fig:BackendFrontendML}. The Data Engineering module is used by the Data Engineer to prepare the dataset for execution in the ML Module. The User Interface module is used by the Data Scientist to specify the architectural configuration, which depends on the context of the problem. The ML Module can be executed manually or with the help of the Autonomic Manager.

The process of manual execution can be visualized in the Sequence Diagram in Figure \ref{fig:DiagramaSequenciaModuloML}. The Data Engineer first prepares the dataset using the Data Engineering module and stores them in the knowledge repository. The Data Scientist then starts the ML Module, passing in the necessary parameters. During execution, the ML Module starts the Monitor component from Autonomic Manager to collect data during execution, retrieves the dataset from the knowledge repository and executes the preprocessing, processing, and post-processing steps defined by the passed parameters to generate the final model. At the end of the execution, the Monitor saves all collected data in the knowledge repository for future use.

\begin{figure*}[h]
\centering
\includegraphics[scale=0.425]{images/Diagrama_Sequencia-manual-eng.jpg}
\caption{Sequence diagram from a manual ML Module execution}
\label{fig:DiagramaSequenciaModuloML}
\end{figure*}

The process of execution with the help of the Autonomic Manager can be visualized in the Sequence Diagram in Figure \ref{fig:DiagramaSequencia}. The Data Scientist, through the User Interface module, selects the dataset and requests an execution. The Autonomic Manager then searches for all previous executions of the ML Module that have already used the modified dataset and are stored in the knowledge repository. It then performs an analysis by performing calculations presented in section~\ref{sec:AnalyzerComponent}. Next, the Autonomic Manager then loops through the configuration of existing execution plans and executes them. It then returns the best execution options to the Data Scientist, who can choose the most appropriate option and request a new execution for the ML Module. This execution is equivalent to the one already explained in Figure \ref{fig:DiagramaSequenciaModuloML}. The Autonomic Manager then searches for the data from this execution to display it to the Data Scientist in the Interface.

\subsection{The ML Module}

The ML module is parameterized to allow the Data Scientist to control the architectural configuration used in its execution. This includes the dataset, protected attribute, and algorithms used. It is implemented using the \textit{Pipes-and-Filters} architecture, which allows for interchangeable Pipes and Filters depending on the parameterization used, making it easy to program execution plans within the Autonomic Manager. These Pipes and Filters are divided into three types of components, illustrated in Figure \ref{fig:BackendFrontendML}, categorized by the newly developed algorithms that were created to improve the learning process's ability to handle biased data:

\begin{itemize}
\item \textbf{Pre-processing:} These algorithms try to remove biased data by transforming it. The implemented algorithms in these components are \textit{Disparate Impact Remover}~\citep{Feldman_2015}, \textit{Learning Fair Representations}~\citep{Zemel_2013}, \textit{Reweighing}~\citep{Kamiran_2011} and \textit{Optimized Preprocessing}~\citep{Calmon_2017}.
\item \textbf{Processing:} These algorithms try to mitigate discrimination during the training process. The implemented algorithms in these components are \textit{Adversarial Debiasing}~\citep{Zhang_2018}, \textit{Exponentiated Gradient Reduction}~\citep{Agarwal_2018}, \textit{Grid Search Reduction}~\citep{Agarwal_2019}, \textit{Meta Fair Classifier}~\citep{Celis_2019}, \textit{Prejudice Remover}~\citep{Kamishima_2012} and \textit{Rich Subgroup Fairness}~\citep{Kearns_2018}.
\item \textbf{Post-processing:} These algorithms modify the predictions of an already trained model to produce fairer outcomes. The implemented algorithms in these components are \textit{Equalized Odds}~\citep{Hardt_2016}, \textit{Calibrated Equalized Odds}~\citep{Pleiss_2017} and \textit{Reject Option Classification}~\citep{Kamiran_2012}.
\end{itemize}

Some steps were additionally implemented in the Pre-processing components: The acquisition of the dataset; acquisition of the protected attribute, privileged group, and unprivileged group; division of the dataset into train, validation and test sets; the data preparation based in the protected attribute, privileged group, and unprivileged group. Because Pre-processing and Post-processing algorithms are not part of model training, Logistic Regression, Gradient Boosting, Random Forest and Support Vector Machines were additionally implemented in the Processing components. The calculation of Evaluation and Fairness metrics was additionally implemented in the Post-processing components.

\subsection{Monitor Component}

The Monitor is initiated along with the execution of the ML Module, as illustrated in Figure \ref{fig:DiagramaSequenciaModuloML}. The following metadata are stored:

\begin{itemize}
\item \textbf{Pre-processing}: Parameters related to the architectural configuration executed by the module (dataset, protected attribute, and pre-processing algorithm) and the checksum of the dataset used.
\item \textbf{Processing}: Parameters related to the architectural configuration executed (training algorithm) and parameters used for training.
\item \textbf{Post-processing}: Parameters related to the architectural configuration executed (post-processing algorithm) and metrics of the resulting model (evaluation metrics and fairness metrics).
\end{itemize}

The data collected per execution are organized into two different sets: One containing the metadata and information related to the architectural configuration executed by the ML Module, and one containing the metrics obtained by the resulting model from that execution. Both sets have an execution identifier to establish consistency between the data in the two sets. This organization facilitates the development of the Analyzer component.

\subsection{Analyzer Component}
\label{sec:AnalyzerComponent}

The Analyzer extracts and analyzes ML Module executions from the knowledge repository based on their metrics, assigning each execution a score that consolidates its evaluation and fairness metrics. The analyses are then grouped by architectural configuration, and each configuration receives a final score. The scores for each architectural configuration are grouped together and placed in a dataset, with additional metadata such as execution date. This dataset is then passed to the Planner to help develop better strategies.

\begin{figure}[h]
\centering
\includegraphics[scale=0.425]{images/metrics-tree.jpg}
\caption {Metrics tree used in execution analysis}
\label{fig:MetricTree}
\end{figure}

To analyze each execution made by the ML Module, the metrics are weighted according to their importance in the context of the problem, as determined by the Data Scientist. This is done to consolidate the evaluation and fairness metrics into a single score, which simplifies planning strategies. This calculation uses a metrics tree, illustrated in Figure \ref{fig:MetricTree}, to divide the metrics into two groups: Evaluation Metrics and Fairness Metrics. The Evaluation Metrics group includes the metrics Accuracy, Precision, Recall, F1-Score, and Area Under the ROC Curve (AUC). The Fairness Metrics group includes the metrics Statistical Parity Difference, or discrimination~\citep{Zemel_2013}, Equal Opportunity Difference~\citep{Biswas_2020}, Average Odds Difference~\citep{Biswas_2020}, Disparate Impact~\citep{Biswas_2020} and Theil Index~\citep{Speicher_2018}. The weights $w_E$ for the Evaluation Metrics group and $w_F$ for the Fairness Metrics group are assigned and the way used to consolidate all metrics was through a score $S$ defined by a weighted mean, displayed in equation \ref{eqn:totalScore}, between the $S_E$ and $S_F$ scores.

\begin{equation}
\label{eqn:totalScore}
	S = \frac{w_F \times S_F + w_E \times S_E}{w_F + w_E}
\end{equation}

To calculate the $S_E$ and $S_F$ for the Evaluation Metrics and Fairness Metrics groups, respectively, a weighted mean is calculated for each group. The weights for each metric are denoted by $w_{m_{Ei}}$ for the Evaluation Metrics group and $w_{m_{Fi}}$ for the Fairness Metrics group, but there are some differences to consider given the nature of the metrics in each group In the case of $S_E$, all calculated metrics are in the interval $\left[0,1 \right]$ and their results are directly proportional, meaning that the best results have the highest values. This implies that a weighted mean can be used without any modifications.

\begin{equation}
\label{eqn:EvaluationScore}
	\begin{aligned}
	S_{E} = \frac{\sum_{i=1}^{n_{m_E}} w_{m_{E_i}} \times m_{E_{i}}}{\sum_{i=1}^{n_{m_E}} w_{m_{E_i}}}
	\end{aligned}
\end{equation}

In the equation \ref{eqn:EvaluationScore}, $n_{m_E}$ is the number of evaluation metrics used (for the tree defined in Figure \ref{fig:MetricTree}, $n_{m_E} = 5$), $w_{m_{E_i}}$ is the weight given to a given evaluation metric and $m_{E_{i}}$ is the value of a given evaluation metric in the analyzed execution.

However, for the case of $S_F$, there are three different intervals and situations for the Fairness metrics present in the group:
